State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-04-28 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
sharpned up the intro of the paper

An estimation framework for neural field models known as dynamical causal modelling (DCM) \cite{David2003} \cite{David2006} has recently been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range connectivity structure between the specific isolated brain regions that best explains a given data set. The interconnected brain regions are modelled by the Jansen and Ritt equations. Another recent publication describing a parameter estimation method with a neural field model used an Unscented Kalman filter with the WC neural field equations \cite{schiff2008kalman}. This work takes a system theoretic approach to the neural estimation problem and marks a first step in what has the potential to revolutionise the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

@@

The work from Schiff and Sauer \cite{schiff2008kalman} successfully demonstrated it is possible to perform state estimation of modified WC equations. This motivates the question, what parameters are the most important? How patient specific do the assumed parameters have to be for the model to be useful? In this paper, we address both of these questions and provide a theoretical platform to perform patient-specific grey-box modelling of the human neocortex. \wtf{not sure about rhetorical questions as a narrative mechanism in general, also not sure how this paper addresses these particular questions.} \omg{Maybe the next section addresses that the connectivity is important. Needs sharpening.}

\section{The Importance of Cortical Structure} Neural field models relate mean firing rates of pre-synaptic neural populations to post-synaptic mean membrane potentials. Each neural population represents a functional cortical processing unit, referred to as a column. The columnar organization of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modeled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity) \cite{Jirsa2009} \cite{Qubbaj2007}. Locally, each column is thought be connected via symmetric short range local excitation, with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field \cite{Sullivan2006}.

Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs the bifurcation points of the system \cite{Hutt2005} and types of oscillations that can be generated \cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to various oscillatory states.

Estimating functional cortical connectivity via EEG promises to shed light on the hierarchy of brain regions that are involved in specific tasks, motivating the use of techniques like DCM. Other comparable methods are based on auto-regressive (AR) modelling of the EEG, MEG, or fMRI times series \todo{[refs!]}. Studies have concentrated on finding functional connectivity patterns using information contained in the AR coefficients using techniques like Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999}. Again, like the DCM approach, these methods estimate long range connectivity patterns that does not provide a clear relationship between the continuum field models and data.

%Analysis of the connectivity estimates provided from these approaches has involved graph theoretical techniques (small worldedness).
Until now, estimation of local intracortical connectivity structure has not been attempted. Recently, it has been shown that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning \cite{Dewar2009}. The key development in this work was to represent the spatiotemporal system as a standard state-space model with an order independent of the number of sensors (ECoG recording electrodes in this case). In addition, the appropriate model selection tools have been developed \cite{Scerri2009} allowing for the application of this technique to neural fields. Modelling the neural dynamics within this framework has the distinct advantage over multivariate AR models, that the number of parameters to define the spatial connectivity is considerably smaller than the number of AR coefficients typically required to achieve the required model complexity. In this paper, we demonstrate for the first time how intracortical connectivity can be inferred from ECoG data, based on a variant of the Wilson and Cowan neural field model \cite{Wilson1973}. This work provides a fundamental link between the theoretical advances in neural field modelling and patient specific data.

%Long range connectivity can be estimated using DTI \cite{Knock2009}, however this short range connectivity can not..
\section{Neural Field Model}\label{NeuralModelSection} In this section, we describe a variant of the WC neural field model \cite{Wilson1973} that was used to create our intracortical connectivity estimator. This model is descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour. 
\begin{todo}
	{Really need to add some citations here.} 
\end{todo}
It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. 
\begin{todo}
	{Same here.} 
\end{todo}
The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology.

The model relates the average number of action potentials $g(\mathbf{r},t)$ arriving at position $r$ to the local post synaptic membrane voltage $v(\mathbf{r},t)$. The post-synaptic potentials generated at a neuronal population at location $\mathbf{r}$ by action potentials arriving from all other connected populations at locations $\mathbf{r}'$ can be described by 
\begin{equation}
	\label{SpikesToPotential} v\left( {\mathbf{r},t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {\mathbf{r},t'} \right)dt'}. 
\end{equation}
The post-synaptic response kernel $h(t)$ is described by \todo{citation needed} \omg{this is about the most vanilla bit} 
\begin{equation}
	\label{SynapticRespKernel} h(t) = \eta(t)\exp{\left(-\zeta t\right)}. 
\end{equation}
where $\zeta=\tau^{-1}$ and $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function. Nonlocal interactions between cortical populations are described by 
\begin{equation}
	\label{RateBasedInteractions} g\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( v\left( \mathbf{r}',t \right) \right)d\mathbf{r}'} + u\left(\mathbf{r},t\right), 
\end{equation}
where $f(\cdot)$ is the firing rate function, $w(\cdot)$ is the spatial connectivity kernel, and $\Omega$ is the spatial domain, representing a cortical sheet or surface. The term $u\left(\mathbf{r},t\right)$ represents a known external input \wtf{Do we actually use the input in this paper, or is it just dragged around the derivation to be discarded in the conclusion?}. The connectivity kernel is typically a ``Mexican hat'' function \todo{which describes strong local activation, weak mid-range repression and weak long-range activation \cite{}}. The exact shape of this kernel is assumed to vary across patients, and hence needs to be inferred from data.

@@

\section{Derivation of Finite Dimensional State-Space Model} In order to implement standard estimation techniques we use a decomposition of the field using a set of Gaussian basis functions defined by

@@

%The method is appropriate for state estimation of stochastic nonlinear dynamical systems. 
\section{Results} The neural field model derived in section~\ref{NeuralModelSection} was used to the generate data for state and parameter estimation. All parameters for the model are given in table~\ref{}. An example of the simulated field can be seen in figure~\ref{}~A. The Figure demonstrates the experimental design where the sensors were placed in the centre of the field (full width at half maximum). This design allowed for uncertainty around the edges of the sensor array, which would be typical from physical data, and alleviated problems associated with boundary conditions. Figure~\ref{}~B. shows example data from the sensors of the model, and figure~\ref{}~C. shows the power spectral density of on of the simulated time-series. The PSD shows the typical $1/f$ characteristics of intracranial EEG.

@@

\section{Discussion}
\begin{itemize}
	\item if we don't know the frequency analysis or the system is marginally sampled or under-sampled, the best we can do is have the basis functions the same as the sensor kernel. \wtf{why wouldn't it be better to have, say, twice as many basis functions as we have sensors, at half the width?} 
	\item alternatives to frequency analysis are model selection... \wtf{This frequency analysis \emph{is} model selection. Maybe what's meant here is that other model selection approaches could be applied? Some sort of IC optimisation?} 
	\item the width of the connectivity basis functions \wtf{do you mean width of the kernel? I'm not sure about this - as the parameters increase the kernel width increases right? So maybe you mean the basis functions, but don't we estimate these widths from the frequency analysis?} here is known exactly. In practice we don't 
	\item the number of basis functions could be set higher to get a better representation of the field, however this is limited by computation complexity. \wtf{Unless you are going to specify the computational complexity, in terms of the number of basis functions, it might be better to say something like ``there is a relationship between the number of basis functions that represent the field and the computational complexity of the estimation algorithm - the more basis functions the longer it takes to estimate''} 
	\item Discuss the novelty of using the reduced model (BrainIDE) for creating patient specific models. \omg{hell yes. This should be front and centre!} 
	\item The size of the dentritic arbour is consistent in the cortex. \omg{?} 
	\item Spatial frequency of field and the relationship to the required density of electrodes/sensors required for estimation or to capture the required information without aliasing. 
	\item The flexibility of the approach 
	\item Combining this method with other to incorporate longer range hetrogeneous connectivity. \omg{awesome} 
	\item The need for activity to perform estimation. \omg{To Learn You Must Perturb!} 
\end{itemize}

\section{Conclusion} 

>>>>

An estimation framework for neural field models known as dynamical causal modelling (DCM) \cite{David2003} \cite{David2006} has recently been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range connectivity structure between the specific isolated brain regions that best explains a given data set. The interconnected brain regions are modelled by the Jansen and Ritt equations. Another recent publication describing a parameter estimation method with a neural field model used an Unscented Kalman filter with the WC neural field equations \cite{schiff2008kalman}. This work takes a system theoretic approach to the neural estimation problem, successfully demonstrating it is possible to perform state estimation of modified WC equations, marking a first step in what has the potential to revolutionise the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

@@

This paper extends the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for higher dimensional systems utilising a systematic model reduction procedure. In addition, a method is presented for estimating the the connectivity structure and the synaptic time constant. Until now, estimation of local intracortical connectivity structure has not been attempted. Recently, it has been shown that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning \cite{Dewar2009}. The key development in this work was to represent the spatiotemporal system as a standard state-space model with an order independent of the number of sensors (recording electrodes in this case). In addition, the appropriate model selection tools have been developed \cite{Scerri2009} allowing for the application of this technique to neural fields. Modelling the neural dynamics within this framework has the distinct advantage over multivariate auto-regressive (AR) models, that the number of parameters to define the spatial connectivity is considerably smaller than the number of AR coefficients typically required to achieve the required model complexity. In this paper, we demonstrate for the first time how intracortical connectivity can be inferred from data, based on a variant of the Wilson and Cowan neural field model \cite{Wilson1973}. This work provides a fundamental link between the theoretical advances in neural field modelling and patient specific data.

The paper proceeds by first deriving the continuum neural field equations in Section~\ref{NeuralModelSection}. Next, a reduced finite dimensional neural field model is derived in Section~\ref{ReducedModelSection}. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. Section~\ref{SpectralAnalysisSection} then establishes conditions using spatial frequency analysis for both sensor and basis function spacing and geometry, such that the dominate dynamics of the neural field can be represented by the reduced model. Following this, the state and parameter estimation procedure is described in Section~\ref{StateAndParameterEstimationSection}. The results for the spatial frequency analysis and parameter estimation are then presented in Section~\ref{ResultsSection}. The implications and limitations of this framework are discussed in Section~\ref{DiscussionSection} along with planned future developments. Finally, we provide concluding remarks in Section~\ref{ConclusionSection}.

\section{Neural Field Model}\label{NeuralModelSection} In this section, we describe a variant of the WC neural field model \cite{Wilson1973} that was used to create our intracortical connectivity estimator. Neural field models relate mean firing rates of pre-synaptic neural populations to mean post-synaptic membrane potentials. They are popular due to being parsimonious yet having a strong link with the underlying physiology. Each neural population represents a functional cortical processing unit, such as a column. The columnar organisation of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modelled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity)~\cite{Jirsa2009}~\cite{Qubbaj2007}. Locally, each column is thought be connected via symmetric short range local excitation, with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}. Neural field models are descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour. 
\begin{todo}
	{Really need to add some citations here.} 
\end{todo}
It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. 
\begin{todo}
	{Same here.} 
\end{todo}

The model relates the average number of action potentials $g(\mathbf{r},t)$ arriving at position $r$ to the local post-synaptic membrane voltage $v(\mathbf{r},t)$. The post-synaptic potentials generated at a neuronal population at location $\mathbf{r}$ by action potentials arriving from all other connected populations at locations $\mathbf{r}'$ can be described by 
\begin{equation}
	\label{SpikesToPotential} v\left( {\mathbf{r},t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {\mathbf{r},t'} \right)dt'}. 
\end{equation}
The post-synaptic response kernel $h(t)$ is described by \todo{citation needed} \omg{this is about the most vanilla bit} 
\begin{equation}
	\label{SynapticRespKernel} h(t) = \eta(t)\exp{\left(-\zeta t\right)}. 
\end{equation}
where $\zeta=\tau^{-1}$ and $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function. Nonlocal interactions between cortical populations are described by 
\begin{equation}
	\label{RateBasedInteractions} g\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( v\left( \mathbf{r}',t \right) \right)d\mathbf{r}'} + u\left(\mathbf{r},t\right), 
\end{equation}
where $f(\cdot)$ is the firing rate function, $w(\cdot)$ is the spatial connectivity kernel, and $\Omega$ is the spatial domain, representing a cortical sheet or surface. The term $u\left(\mathbf{r},t\right)$ represents a known external input \wtf{Do we actually use the input in this paper, or is it just dragged around the derivation to be discarded in the conclusion?}. The connectivity kernel is typically a ``Mexican hat'' function \todo{which describes strong local activation, weak mid-range repression and weak long-range activation \cite{}}. The exact shape of this kernel is assumed to vary across patients, and hence needs to be inferred from data.

@@

\section{Derivation of Finite Dimensional State-Space Model}\label{ReducedModelSection} In order to implement standard estimation techniques we use a decomposition of the field using a set of Gaussian basis functions defined by

@@

%The method is appropriate for state estimation of stochastic nonlinear dynamical systems. 
\section{Results}\label{ResultsSection} The neural field model derived in section~\ref{NeuralModelSection} was used to the generate data for state and parameter estimation. All parameters for the model are given in table~\ref{}. An example of the simulated field can be seen in figure~\ref{}~A. The Figure demonstrates the experimental design where the sensors were placed in the centre of the field (full width at half maximum). This design allowed for uncertainty around the edges of the sensor array, which would be typical from physical data, and alleviated problems associated with boundary conditions. Figure~\ref{}~B. shows example data from the sensors of the model, and figure~\ref{}~C. shows the power spectral density of on of the simulated time-series. The PSD shows the typical $1/f$ characteristics of intracranial EEG.

@@

\section{Discussion}\label{DiscussionSection}
\begin{itemize}
	\item if we don't know the frequency analysis or the system is marginally sampled or under-sampled, the best we can do is have the basis functions the same as the sensor kernel. \wtf{why wouldn't it be better to have, say, twice as many basis functions as we have sensors, at half the width?} 
	\item alternatives to frequency analysis are model selection... \wtf{This frequency analysis \emph{is} model selection. Maybe what's meant here is that other model selection approaches could be applied? Some sort of IC optimisation?} 
	\item the width of the connectivity basis functions \wtf{do you mean width of the kernel? I'm not sure about this - as the parameters increase the kernel width increases right? So maybe you mean the basis functions, but don't we estimate these widths from the frequency analysis?} here is known exactly. In practice we don't 
	\item the number of basis functions could be set higher to get a better representation of the field, however this is limited by computation complexity. \wtf{Unless you are going to specify the computational complexity, in terms of the number of basis functions, it might be better to say something like ``there is a relationship between the number of basis functions that represent the field and the computational complexity of the estimation algorithm - the more basis functions the longer it takes to estimate''} 
	\item Discuss the novelty of using the reduced model (BrainIDE) for creating patient specific models. \omg{hell yes. This should be front and centre!} 
	\item The size of the dentritic arbour is consistent in the cortex. \omg{?} 
	\item Spatial frequency of field and the relationship to the required density of electrodes/sensors required for estimation or to capture the required information without aliasing. 
	\item The flexibility of the approach 
	\item Combining this method with other to incorporate longer range hetrogeneous connectivity. \omg{awesome} 
	\item The need for activity to perform estimation. \omg{To Learn You Must Perturb!} 
	\item Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs the bifurcation points of the system \cite{Hutt2005} and types of oscillations that can be generated \cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to various oscillatory states.
	\item Estimating functional cortical connectivity via EEG promises to shed light on the hierarchy of brain regions that are involved in specific tasks, motivating the use of techniques like DCM. Other comparable methods are based on auto-regressive (AR) modelling of the EEG, MEG, or fMRI times series \todo{[refs!]}. Studies have concentrated on finding functional connectivity patterns using information contained in the AR coefficients using techniques like Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999}. Again, like the DCM approach, these methods estimate long range connectivity patterns that does not provide a clear relationship between the continuum field models and data.
	\item Long range connectivity can be estimated using DTI \cite{Knock2009}, however this short range connectivity can not estimated using DTI
\end{itemize}

\section{Conclusion}\label{ConclusionSection}
